\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb,latexsym,amsmath,epsfig,amsthm}
\usepackage{enumitem}
\usepackage{ulem}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{makecell}


\title{Homework 2}
\author{Lucien Gaitskell}
\date{December 2020}

\begin{document}

\maketitle

\section{Problem 1}

$S = [34,32,21,9,43,35,42,15,1,4,17,20,8,42,27]$

\begin{itemize}
    \item $\textproc{DeterministicSelect}(S, 8)$ \\
    Baby sequences: $[[34,32,21,9,43],[35,42,15,1,4],[17,20,8,42,27]]$ \\
    Baby medians: $[32,15,20]$

    \begin{itemize}
        \item $\textproc{DeterministicSelect}([32, 15, 20],2)$ \\
        Baby sequences: $[[32,15,20]]$ \\
        Baby medians: $[20]$

        \begin{itemize}
            \item $\textproc{DeterministicSelect}([20],1)$ \\
            return $20$

        \end{itemize}
        $x = 20 $ \\
        $L = [15] $ \\
        $E = [20] $ \\
        $G = [32] $ \\
        return $20$

    \end{itemize}

    $x = 20$ \\
    $L = [1,4,8,9,15,17]$ \\
    $E = [20]$ \\
    $G = [34,32,21,43,35,42,42,27]$

    \begin{itemize}
        \item $\textproc{DeterministicSelect}([34,32,21,43,35,42,42,27],1)$ \\
        Baby sequences: $[[34,32,21,43,35],[32,42,42,27]]$ \\
        Baby median: $[34, 42]$
        \begin{itemize}
            \item $\textproc{DeterministicSelect}([34, 42],1)$ \\
            Baby sequences: $[[34, 42]]$ \\
            Baby medians: $[34]$
            \begin{itemize}
                \item $\textproc{DeterministicSelect}([34],1)$ \\
                return $34$
            \end{itemize}
            
            $x = 34$ \\
            $L = []$ \\
            $E = [34]$ \\
            $G = [42]$
            
            return $34$
        \end{itemize}

        $x = 34$ \\
        $L = [21, 32, 27]$ \\
        $E = [34]$ \\
        $G = [43, 35, 42, 42]$

        \begin{itemize}
            \item $\textproc{DeterministicSelect}([21, 32, 27],1)$ \\
            Baby sequences: $[[21, 32, 27]]$ \\
            Baby medians: $[27]$
            \begin{itemize}
                \item $\textproc{DeterministicSelect}([27],1)$ \\
                return $27$
            \end{itemize}

            $x = 27$ \\
            $L = [21]$ \\
            $E = [27]$ \\
            $G = [32]$

            \begin{itemize}
                \item $\textproc{DeterministicSelect}([21], 1)$ \\
                return $21$
            \end{itemize}
            return $21$           
        \end{itemize}
        
        return $21$
    \end{itemize}

    return $21$
\end{itemize}


\pagebreak
\section{Problem 2}

\subsection{Part 1}

Assume the day begins at $d_o$ and ends as $d_e$.

\begin{enumerate}[label=(\alph*)]
    \item This is not an optimal algorithm.

    \textbf{Counterexample:} Suppose workshop $1$, such that $s_1 = d_o$
    and $t_1 = d_e$. Suppose there is another workshop ($2$) that starts a minute
    after workshop $1$ and ends during workshop $1$. $\therefore s_2=s_1 + 1, t_2=x$
    such that $s_1+1 < x < t_1$. Suppose another workshop $3$, such that 
    $s_3 = x, t_3 = d_e$.

    This greedy algorithm will select workshop $1$ due to the earlier starting time.
    However, selecting workshop $1$ will cause workshop $2$ and $3$ to be removed.
    Therefore, only one workshop will be selected to visit,
    despite the opportunity to visit two during the day.

    \item This is an optimal algorithm.
    
    \textbf{Proof:} Let each workshop be $w_i$ in $W$.
    For each workshop $w_i$ there is a set $O_i$ of the workshops overlapping $w_i$.
    Let $w_a$ and $w_b$ be overlapping workshops,
    with $w_a$ having the latest start time in $W$.

    Therefore, it is not possible for another workshop to begin after $w_a$.
    Thus, as $w_b$ overlaps with $w_a$, $O_a \subset O_b$.
    Given this situation, $|O_a| \leq |O_b|$ must be true for any $w_b$ overlapping $w_a$.
    Therefore, it must be optimal to select $w_a$.

    Next, $w_a$ and all $w_b$ are removed from the available workshops,
    and the process begins once again.
    The new $w_a$ will be the remaining workshop with the latest starting time.

    \item The is not an optimal algorithm.
    
    \textbf{Counterexample:} Suppose three workshops:
        \begin{enumerate}[label=\arabic*]
            \item Length x
            \item Length x+1
            \item Length x+1
        \end{enumerate}

        Suppose workshop $2$ and $3$ immediately follow each other, and workshop $1$ overlaps both.
        Workshop $1$ will be selected first, and others will be excluded.
        However, selecting workshop $2$ and $3$ is the more optimal solution.
\end{enumerate}

\subsection{Part 2}

The pseudocode can be found in \textbf{Algorithm \ref{standsalgo}}

\begin{algorithm}
    \caption{Workshop visits}
    \label{standsalgo}
    \begin{algorithmic}
        \Function{VisitStands}{$S$} \Comment{given $S$ the list of stands}
            \State $V \gets []$     \Comment{Stores each visit}
            \State $S_{sorted} \gets \Call{SortEarliestEnd}{S}$
                \Comment{Sort in earliest to latest end time}
            \item[]
            \For{$w$ in $S_{sorted}$}
                \State Remove $w$ from $S_{sorted}$
                \State $G \gets [w]$
                \item[]
                \State Get all workshops $A$ overlapping with w
                \State Remove all $A$ from $S_{sorted}$
                \State Add $A$ to $G$
                \item[]
                \State Add $G$ to $V$
            \EndFor

            % PROBLEM: This doesn't select times
            \State \Return $V$
        \EndFunction
    \end{algorithmic}
\end{algorithm}

This algorithm must return the optimal solution.
It is proved by induction... \\
\textbf{I need to learn more rigorous proofs}

\pagebreak
\section{Problem 3}

\begin{enumerate}[label=\alph*)]
    \item There are only three possible groupings of staff members within 10 miles of each other.
        It is important to note that two of these groups share a member.
        Regardless, there will need to be three stations in a configuration such as:
        \begin{enumerate}[label={Station \arabic*:}, leftmargin=*]
            \item 5
            \item 15
            \item 30
        \end{enumerate}
    \item Process:
        The pseudocode is documented in \textbf{Algorithm \ref{wifialgo}}
        \begin{algorithm}
            \caption{WiFi towers}
            \label{wifialgo}
            \begin{algorithmic}[1]
                \Function{WiFiTowers}{$N$}  \Comment{given $N$ the location of customers}
                    \State Let $T$ be the an empty set of towers
                    \item[]
                    \State Let $N_{sorted}$ be $N$ in increasing distance along $l$
                    \For{$c_i$ in $N_{sorted}$}
                        \State Remove $c_i$ from $N_{sorted}$
                        \State Let $p_i$ be the position along $l$ of $c_i$
                        \item[]
                        \State Create new tower $t_i$ of position $p_i + d/2$
                        \State Add new tower $t_i$ to $T$
                        \item[]
                        \For{$c_j$ in $N_{sorted}$}
                            \If{$c_j$ overlaps with $t_i$}
                                \State Remove $c_j$ from $N_{sorted}$
                            \Else
                                \State Break from inner loop
                            \EndIf
                        \EndFor
                    \EndFor
                \EndFunction
            \end{algorithmic}
        \end{algorithm}
        
        By first sorting the set, and then beginning from the smaller side,
        the station must be $\geq x_1$ for $x_i$ the position of next smallest
        distance staff $i$.
        Therefore, all the other customers within $d$ to the right can share a station.
        In subsequent iterations, all customers to the left have already been
        handled by a tower, and therefore no longer need to be considered.
        This continues until all customers members are visited.
        This must be the most efficient, as as a bare minimum, every staff member
        requires a station, however can be grouped to save on overall count.

        The runtime will be $O(n \log{n} + n) = O(n \log{n})$.
        This is because a sort is first necessary ($O(n \log{n})$).
        Subsequently, the $n$ customers are grouped.
        As each customer is visited only once, the loop takes $O(n)$ time.

    \item In this case, the consumers will still be sorted by distance along $\ell$.
        The search will have to be performed until the next consumer
        has $\Delta \ell > d$.
        However, not all consumers within this range will be grouped.
        A consumer will be in range and will be grouped if their distance
        $d \leq \sqrt{(\Delta x- \frac{d}{2})^2 + y^2}$.
        The grouped consumers will be removed from $N$, and then the algorithm
        will start again at the beginning of the remainin sorted set.
        
    \item Yes, this algorithm does depend on a linear arrangement of staff.
        For example, if a all the staff are arranged on a spiral, beginning at an origin
        and with a  radius always smaller than $5$. The optimal solution is a single station
        at the origin. However, the algorithm would only consider along the line, and would
        add multiple stations.
\end{enumerate}

\pagebreak
\section{Problem 4}

\begin{enumerate}
    \item $T(i,j)$
        \textbf{Base cases}:
        \begin{enumerate}
            \item if $i = 0 \therefore \text{return } j $ representing length of $Y[1..j]$
            \item if $j = 0 \therefore \text{return } i $ representing length of $X[1..i]$
        \end{enumerate}

        if last characters match: $X[i] = Y[j] = A \therefore subcost = 0 \text{ else } subcost = 1$

        \textbf{Recursive step} - return minimum of:
        \begin{enumerate}
            \item $T(i-1, j-1) + subcost$ (substitution, if necessary)
            \item $T(i-1, j) + 1$ (delete from X)
            \item $T(i, j-1) + 1$ (insertion as not in X)
        \end{enumerate}
    
        The base cases are inherently true, as only insertions can occur for (a)
        and deletions for (b).
        Each recursive step represents the three possible operations for each sub-string,
        as the values in the string can only be inserted, deleted, or substituted.
        
        Therefore, as each step and the base cases are true, the the algorithm holds.

    \item \textbf{$T$ Table: }\\
        \begin{tabular}{c||c|c|c|c|c|c|c|c|c|}
            & \thead{0} & \thead{P\\1} & \thead{O\\2} & \thead{R\\3} & \thead{K\\4} & \thead{C\\5} & \thead{H\\6} & \thead{O\\7} & \thead{P\\8} \\
            \hline
            \hline
            0   & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 \\
            \hline
            T 1 & 1 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 \\
            \hline
            O 2 & 2 & 2 & 1 & 2 & 3 & 4 & 5 & 6 & 7 \\
            \hline
            N 3 & 3 & 3 & 2 & 2 & 3 & 4 & 5 & 6 & 7 \\
            \hline
            K 4 & 4 & 4 & 3 & 3 & 2 & 3 & 4 & 5 & 6 \\
            \hline
            A 5 & 5 & 5 & 4 & 4 & 3 & 3 & 4 & 5 & 6 \\
            \hline
            T 6 & 6 & 6 & 5 & 5 & 4 & 4 & 4 & 5 & 6 \\
            \hline
            S 7 & 7 & 7 & 6 & 6 & 5 & 5 & 5 & 5 & 6 \\
            \hline
            U 8 & 8 & 8 & 7 & 7 & 6 & 6 & 6 & 6 & \textbf{6} \\
            \hline
        \end{tabular}

        I first filled out the row along T/$1$ and column below P/$1$,
        following the base cases where $i = 0$ and $j = 0$.
        Subsequently, I computed the other cells starting at $i,j=(1,1)$,
        requiring that the three relative cells to the top, top left, and left were also filled.
        I propagated down each subsequent column and row until reaching $i,j=(8,8)$.
    
    
    \item The issue will arise if the two strings are: \emph{ABCDE} and \emph{FGAHI}.
        This is because the altered value will become an issue, as $T(1, 3) = 3$ and $\not= 2$
        as a substition is taking place. However, the algorithm wll instead favor a deletion.
        Otherwise, if $X[1]\not=Y[3]$, then the initializaton error would not cause an issue,
        as deletion should be favored.

\end{enumerate}

\section{Problem 5}

\begin{enumerate}[label=\alph*.]
    \item Algorithm \ref{marginalgo} given $S$ an array of words in the sentence

    \begin{algorithm}
        \caption{Margin penalty calculation}
        \label{marginalgo}
        \begin{algorithmic}[1]
            \Function{Margin}{$S, m$} \Comment{Cost for sentence $S$ in margin size $m$}
                \If{$isempty(S)$}
                    \State \textbf{return} 0
                \EndIf
                \item[]
                \State Create array $C[1\dots n][1\dots n]$ \Comment{contains total characters of words $i\to j$}
                \State Create array $T[1\dots n+1]$         \Comment{cost of optimal arrangement of words $1\to i$}
                \State Create arary $P[1\dots n+1]$         \Comment{word at the start of last line given $i$ words}
                \State $T[0]=0$ and $P[0]=0$
                \State Create empty list $b$                \Comment{final list of words starting sentences}
                \item[]
                \For{$i=0\to n-1$}
                    \State $C(i,i) = len(i)$
                    \For{$j=i\to n-1$}
                        \State $C(i,j+1)=C(i,j) + 1 + len(j+1)$
                    \EndFor
                \EndFor
                \item[]
                \For{$j=1\to n$}    \Comment{determine optimal cost if $j$ ends a line}
                    \State $T(j) = \min_{i \in (0\dots j-1)}{\lbrack T(i) + (m- C(i,j-1))^2 \rbrack}$
                    \item[]         \Comment{try each $i$ ending previous line}
                    \State $P(j) = i$ from above that minimized value
                \EndFor
                \item[]
                \State $w=\min{T(i)}$ such that $\{i, C(i, n-1) \leq m\}$
                \While{$w\not=0$}   \Comment{track back to build up newlines list}
                    \State Add $w$ to $b$   \Comment{add to newlines list}
                    \State $w = P(w)$
                \EndWhile
                \item[]
                \State\Return $b$
            \EndFunction
        \end{algorithmic}
    \end{algorithm}

    \item This dynamic programming algorithm calculate the post for all different sets of sequential words.
    Therefore, as the minimum sets are used, it must return the optimal solution.
    \item The runtime will be $O(n^2) + O(n^2) + O(n) = O(n^2)$ for number of words $n$,
    as there are two pairs of two nested loops operating along the length of the set, in series,
    along with a final single loop.
\end{enumerate}

\end{document}
